---
title: "`r params$title`"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  dataset: NULL
  title: "Single-cell RNA-seq Analysis: Preliminary Results"
---

<!--
Render this report for one analysis group:

    Rscript -e "rmarkdown::render('08_report.Rmd', params = list(dataset = 'TISSUE_A'))"

or on a cluster:

    sbatch ../../integrated_analysis/R/render-report.sbatch 08_report.Rmd

Leaving params$dataset as NULL uses the first group in config/samples.tsv.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 8)

library(Seurat)
library(dplyr)
library(ggplot2)
library(readr)
library(knitr)
library(patchwork)
library(stringr)

# Find and load the shared config by walking up to the repo root.
.p <- normalizePath(getwd(), mustWork = FALSE)
while (!file.exists(file.path(.p, "02_R", "R", "config.R")) && dirname(.p) != .p) {
  .p <- dirname(.p)
}
if (!file.exists(file.path(.p, "02_R", "R", "config.R"))) {
  stop("Could not find 02_R/R/config.R. Knit this report from inside the repository.")
}

# The dataset parameter, when supplied, selects which group to report on.
if (!is.null(params$dataset)) DATASET <- params$dataset
source(file.path(.p, "02_R", "R", "config.R"))

obj             <- readRDS(file.path(OUTPUT_DIR, "objects", "06_final_analysis.rds"))
sample_meta     <- readRDS(file.path(OUTPUT_DIR, "objects", "01_sample_metadata.rds"))
cluster_markers <- read_csv(file.path(OUTPUT_DIR, "tables", "05_cluster_markers.csv"),
                            show_col_types = FALSE)
qc_summary      <- read_csv(file.path(OUTPUT_DIR, "tables", "02_qc_summary_before.csv"),
                            show_col_types = FALSE)

vdj_path <- file.path(OUTPUT_DIR, "tables", "03_vdj_summary.csv")
vdj_summary <- if (file.exists(vdj_path)) read_csv(vdj_path, show_col_types = FALSE) else NULL

n_cells      <- ncol(obj)
n_clusters   <- length(unique(obj$seurat_clusters))
n_samples    <- nrow(sample_meta)
conditions   <- sort(unique(obj$condition))
cluster_var  <- if ("cell_type" %in% colnames(obj@meta.data)) "cell_type" else "seurat_clusters"
publishing   <- nzchar(CFG$WEB_PUBLISH_URL %||% "") && !grepl("example\\.org", CFG$WEB_PUBLISH_URL)
```

# Executive summary

Analysis group: **`r DATASET`**
Conditions compared: **`r paste(conditions, collapse = " vs ")`**

- `r n_samples` samples, `r format(n_cells, big.mark = ",")` high-quality cells after filtering
- `r n_clusters` transcriptional clusters identified
```{r vdj-bullets, results='asis'}
if (sum(obj$has_tcr, na.rm = TRUE) > 0) {
  cat("- TCR detected in ", sum(obj$has_tcr, na.rm = TRUE), " cells (",
      round(100 * sum(obj$has_tcr, na.rm = TRUE) / n_cells, 1), "%)\n", sep = "")
}
if (sum(obj$has_bcr, na.rm = TRUE) > 0) {
  cat("- BCR detected in ", sum(obj$has_bcr, na.rm = TRUE), " cells (",
      round(100 * sum(obj$has_bcr, na.rm = TRUE) / n_cells, 1), "%)\n", sep = "")
}
```

<!-- EDIT: replace with the actual biological findings for your study. -->

# Methods

**Preprocessing.** Libraries were processed with Cell Ranger
(`r CFG$CELLRANGER_MODULE`) using `cellranger multi`, aligning to the
`r CFG$SPECIES` reference. Libraries sequenced:
`r paste(LIBRARY_TYPES, collapse = ", ")`.
`r if (CFG$USE_HASHTAG_DEMUX == "true") "Pooled specimens were demultiplexed using hashtag antibodies." else ""`

**Quality control.** Cells were retained based on the number of detected genes,
total UMI count, and mitochondrial and haemoglobin content. Exact thresholds are
recorded at the top of `02_qc_filtering.R`.

**Normalisation and clustering.** Counts were log-normalised (scale factor
10,000), the top variable genes selected by variance-stabilising transformation,
and principal components computed. Graph-based clustering and UMAP embedding
used the leading principal components; parameters are at the top of
`04_norm_and_cluster.R`.

```{r methods-vdj, results='asis'}
if (has_library("TCR") || has_library("BCR")) {
  cat("**V(D)J integration.** Only high-confidence, productive contigs from real",
      "cells were retained and summarised per barcode before being joined to the",
      "gene expression object.\n\n")
}
```

**Differential expression.** Conditions were compared overall and within each
cluster. Note that per-cell tests treat cells as independent replicates and
therefore overstate significance; treat these results as exploratory and
confirm important findings with a pseudobulk analysis across biological
replicates.

# Results

## Data quality

```{r qc-summary}
kable(
  qc_summary %>%
    mutate(
      median_genes = format(median_genes, big.mark = ","),
      median_umis  = format(median_umis, big.mark = ","),
      median_mt    = paste0(round(median_mt, 2), "%"),
      median_ribo  = paste0(round(median_ribo, 2), "%")
    ),
  caption = "Quality control metrics by sample (before filtering)"
)
```

```{r web-summaries, results='asis'}
if (publishing) {
  cat("**Cell Ranger web summaries:**\n\n")
  for (sid in sample_meta$sample_id) {
    cat("- [", sid, "](", WEB_URL, "/web_summaries/", sid, "_web_summary.html)\n", sep = "")
  }
}
```

## Clustering

```{r umap, fig.height=6}
DimPlot(obj, group.by = cluster_var, label = TRUE, label.size = 3, repel = TRUE) +
  ggtitle(paste0(n_clusters, " clusters identified")) +
  theme(plot.title = element_text(size = 14, hjust = 0.5))
```

```{r cluster-sizes}
sizes <- as.data.frame(table(obj@meta.data[[cluster_var]]),
                       stringsAsFactors = FALSE)
colnames(sizes) <- c("Cluster", "Cells")
sizes$Percentage <- round(100 * sizes$Cells / n_cells, 1)
kable(sizes, caption = "Cells per cluster")
```

```{r top-markers}
kable(
  cluster_markers %>%
    group_by(cluster) %>%
    slice_max(n = 3, order_by = avg_log2FC) %>%
    select(cluster, gene, avg_log2FC, p_val_adj),
  caption = "Top 3 marker genes per cluster",
  digits = 3
)
```

## Composition by condition

```{r composition, fig.height=6}
DimPlot(obj, group.by = "condition") + ggtitle("Cells by condition")
```

```{r composition-table}
comp <- as.data.frame.matrix(table(obj@meta.data[[cluster_var]], obj$condition))
comp$Cluster <- rownames(comp)
comp$Total <- rowSums(comp[, conditions, drop = FALSE])
for (cond in conditions) {
  comp[[paste0(cond, " %")]] <- round(100 * comp[[cond]] / comp$Total, 1)
}
kable(comp[, c("Cluster", conditions, paste0(conditions, " %")), drop = FALSE],
      caption = "Cluster composition by condition", row.names = FALSE)
```

```{r vdj-results}
if (!is.null(vdj_summary)) {
  kable(vdj_summary, caption = "V(D)J detection by sample")
}
```

## Differential expression

```{r de-summary, results='asis'}
de_files <- list.files(file.path(OUTPUT_DIR, "tables"),
                       pattern = "^06_de_.*\\.csv$", full.names = TRUE)

if (length(de_files) == 0) {
  cat("No differential expression results found. Run `06_de.R` first.\n")
} else {
  de_summary <- do.call(rbind, lapply(de_files, function(f) {
    d <- read_csv(f, show_col_types = FALSE)
    data.frame(
      Comparison  = sub("^06_de_(overall_)?", "", tools::file_path_sans_ext(basename(f))),
      Genes       = nrow(d),
      Significant = sum(d$p_val_adj < 0.05, na.rm = TRUE),
      Link        = if (publishing) {
        paste0("[Download](", WEB_URL, "/tables/", basename(f), ")")
      } else {
        basename(f)
      },
      stringsAsFactors = FALSE
    )
  }))
  print(kable(de_summary[order(-de_summary$Significant), ],
              caption = "Differential expression results",
              row.names = FALSE, escape = FALSE))
}
```

```{r downloads, results='asis'}
if (publishing) {
  cat("# Downloads\n\n")
  cat("- [All plots](", WEB_URL, "/plots/)\n", sep = "")
  cat("- [All tables](", WEB_URL, "/tables/)\n", sep = "")
  cat("- [Cell Ranger reports](", WEB_URL, "/web_summaries/)\n", sep = "")
}
```

# Session information

```{r session-info}
sessionInfo()
```
